Causal inference and the detection of coevolutionary contingencies using generalized dynamic phylogenetic models

### 2024-09-05 manuscript.Rmd
Update manuscript.Rmd
finalized abstract, spelled out conclusions and limitations

\begin{center}
Phylogenetic comparative methods are widely used to study the coevolution of biological or cultural traits over evolutionary time. Among the most popular of these methods are phylogenetic generalised linear mixed models, phylogenetic path analysis, Ornstein-Uhlenbeck models, and Pagel's `discrete' method. While these methods are appropriate under specific scenarios, they also limited by often unrealistic assumptions about the evolutionary process and limits on the  number and type of traits that researchers can include. Here, we develop a novel class of \textit{generalized dynamic phylogenetic models} (GDPMs) that overcomes these limitations and synthesizes the strengths of existing methods into a generalizable framework for dynamic phylogenetic inference. Treating the phylogeny as an implicit time series, GDPMs model the co-evolution of any number of traits undergoing both selection and drift, capable of inferring directed evolution (X -> Y vs Y -> X), feedback (X <-> Y), and contigencies (e.g., first X, then Y). We introduce the \texttt{coevolve} R package, a user-friendly package purpose-built for fitting GDPMs in a Bayesian framework using Stan. We provide a short tutorial for fitting these models and present several empirical applications of GDPMs to make inferences about trait coevolution, including analyses of brain size in non-human primates and social complexity in human societies. ADD BRIEF SUMMARY/HIGHLIGHTS OF THE EMPIRICAL RESULTS.
\end{center}

@@

# Conclusions and Limitations

We presented 
First paragraph repeats major advances of the method:
Dynamic model, analogous to Pagel’s discrete, but without being limited to binary traits
Can handle any distribution, latent factors
Can detect evolutionary contingencies (A→B, B→A) and reciprocal causation
Unbiased estimator, empirical examples summary

Second paragraph states limitations, starting with specific to this method then going broader
1: Correlated drift cannot be separated from cross-selection (though this is possible with ‘real’ longitudinal data) - biological examples of traits under correlated drift?
2: Cross-selection effects assumed to be linear - this can be somewhat relaxed with clade-specific parameters, but still locally linear
3: Computational cost - as with any Bayesian model - can range from minutes (for small phylogenies / datasets) to hours (SCCS example) or even days (for the most complex models, or if looped over a sample of trees)
4: Interpretation - we’re detecting contingency, not true causality; though this is the same for other kinds of causal inference (Pearl, Pagel, Turchin)
5: Data quality and inference - comparative data tend to be crude, and all comparative analyses assume that extant variation reflects all past variation. Virtual archeology / paleontology is not as good as real archeology / paleontology! This can be somewhat amended by including information about ancestral nodes, e.g. complexity SCCS analysis put prior for low complexity on last common ancestor - in principle this could be done for any internal node, with prior reflecting exact state of knowledge

>>>>

\begin{center}
Phylogenetic comparative methods are widely used to study the coevolution of biological or cultural traits over evolutionary time. Among the most popular of these methods are phylogenetic generalised linear mixed models, phylogenetic path analysis, Ornstein-Uhlenbeck models, and Pagel's `discrete' method. While these methods are appropriate under specific scenarios, they also limited by often unrealistic assumptions about the evolutionary process and limits on the  number and type of traits that researchers can include. Here, we develop a novel class of \textit{generalized dynamic phylogenetic models} (GDPMs) that overcomes these limitations and synthesizes the strengths of existing methods into a generalizable framework for dynamic phylogenetic inference. Treating the phylogeny as an implicit time series, GDPMs model the co-evolution of any number of traits undergoing both selection and drift, capable of inferring directed evolution (X -> Y vs Y -> X), feedback (X <-> Y), and contigencies (e.g., first X, then Y). We introduce the \texttt{coevolve} R package, a user-friendly package purpose-built for fitting GDPMs in a Bayesian framework using Stan, accompanied by a short tutorial. We use synthetic data and simulation-based calibration to show that our models can recover true parameter values and yield unbiased estimates. We also present several empirical applications of GDPMs, including analyses of brain size, life history and socio-ecology in non-human primates and social complexity in human societies. These examples highlight the flexibility of the Bayesian implementation, allowing researchers to model latent variables, mulitlevel structures and repeated measures, measurement error, missing data, and other complexities inherent in comparative data.
\end{center}

@@

# Conclusions and Limitations (ROUGH DRAFT!)

We presented a novel class of phylogenetic methods that allows the detection of coevolutionary dynamics and has several advantages over existing approaches. These include an explicitly dynamic approach, which can detect evolutionary contingencies and reciprocal causation, coupled with the enormous flexibility of the Bayesian implementation in the Stan programming language, allowing researchers to naturally account for latent factors, repeated measures, measurement error, missing data, and other real-life complexities of comparative data. The coevolve R package supports some of these complexities (such as missing data), as well as a reasonable diversity of distributions, facilitating the adoption of this method for most R users, while some researchers might prefer the flexibility of coding their models directly in Stan.

Despite several advances, our method is still subject to several limitations, some specific to GDPMs, and some more broadly applicable. In cases where traits experience correlated drift, GDPMs will struggle to separate this effect from cross-selection (EXPAND ON THIS, PERHAPS WITH A REAL BIOLOGICAL EXAMPLE?). GDPMs assume that cross-selection effects are linear; this assumption can be somewhat relaxed by using clade-specific parameters, though the effects will still be 'locally' linear (i.e. within a given clade). As with any Bayesian estimation, GDPMs can be computationally costly, though we note that the Hamiltonian Monte Carlo sampler of Stan is substantially more efficient than conventional MCMC samplers, typically requiring only a few thousand instead of millions of iterations; the models presented in this paper generally took minutes to hours to run. As with all comparative analyses, GDPMs are subject to concerns about data quality and bias; e.g. that extant trait variation reflects past variation, that available phylogenies accurately reflect true evolutionary history, etc. We also note that since GDPMs work as a time series, inference or knowledge about ancestral states is particularly crucial, and relevant information should be incorporated into the models - for instance, Ringen et al [@Ringen_preprint] used priors for low complexity on the last common ancestor of all human societies; in this way, virtual archeology (or paleontology) can be enhanced by real archeology. Lastly, we caution that GDPMs detect contingencies, and not causation per se, though this caveat applies to all approaches to causal inference from observational data [@Pearl2009; @Pagel1994].
